---
title: "Forecasting the 2026 Ebola Outbreak"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecasting the 2026 Ebola Outbreak}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 7,
  dpi = 300
)
```

## Overview

In May 2026, the Democratic Republic of the Congo reported an Ebola outbreak
caused by the Bundibugyo strain. By July, 2,423 cases and 967 deaths had been
reported. This example forecasts incidence using surveillance data from the
[Institut National de Recherche Biomédicale](https://github.com/INRB-UMIE/BDBV2026-Data).

The vignette loads precomputed data, cross-validation and forecast objects.
Model-fitting chunks are disabled, but all plots are generated when the article
is built.

## Data

The example uses the six locations with the most confirmed cases. Missing days
carry the previous cumulative count forwards, and downward revisions are
removed. Leading zeros are dropped before log transformation.

```{r, data, eval = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

ebola <- read.csv(
  "https://raw.githubusercontent.com/scc-usc/ebola2026/refs/heads/main/hubverse_observed_data.csv"
) |>
  filter(
    target == "insp_sitrep__cumulative_confirmed_cases__daily",
    !is.na(location)
  ) |>
  mutate(target_end_date = as.Date(target_end_date))

top_locations <- ebola |>
  slice_max(target_end_date, by = location) |>
  slice_max(observation, n = 6) |>
  pull(location)

ebola <- ebola |>
  filter(location %in% top_locations) |>
  arrange(location, target_end_date) |>
  group_by(location) |>
  complete(
    target_end_date = seq(
      min(target_end_date),
      max(target_end_date),
      by = "day"
    )
  ) |>
  fill(observation, .direction = "down") |>
  mutate(observation = cummax(coalesce(observation, 0))) |>
  filter(cumsum(observation > 0) > 0) |>
  ungroup() |>
  mutate(target = "insp_sitrep__cumulative_confirmed_cases__daily")
```


## `incast` workflow

### Validate the data

`check_data()` standardises the columns and returns an `incast_data` object.

```{r, check_data, eval = FALSE}
library(incast)
data <- ebola |> check_data()
```

```{r, load-data, include = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(incast)
data <- readRDS("ebola_data.rds")
```

```{r, show-data, fig.cap = "Daily cumulative confirmed cases by location."}
data
data |> autoplot()
```

Revision history is unavailable, so the workflow skips `get_ncast()`.


### Cross-validation

Compare the default models with a custom ARIMA model, a neural network,
Prophet and four foundation models:

```{r, models, eval = FALSE}
library(fable)
library(fable.prophet)

models <- c(
  default_models(),
  list(
    CUSTOM_ARIMA = ARIMA(log(observation) ~ pdq(1, 1, 0)),
    NNETAR = NNETAR(log(observation), n_networks = 10),
    PROPHET = prophet(log(observation)),
    CHRONOS = FOUNDATION(log(observation), "chronos"),
    TIMESFM = FOUNDATION(log(observation), "timesfm"),
    SUNDIAL = FOUNDATION(log(observation), "sundial"),
    MOIRAI = FOUNDATION(log(observation), "moirai")
  )
)
```

Forecast seven days ahead from 16 daily origins. The evaluation covers 22 days.

```{r, cross-validation, eval = FALSE}
cv <- data |>
  get_cv(
    h = 7,
    step = 1,
    n_origins = 16,
    models = models
  )
```

```{r, load-cv, include = FALSE}
cv <- readRDS("ebola_cv.rds")
```

```{r, show-cv, fig.cap = "Cross-validation performance by model and location."}
cv
cv |>
  autoplot() +
  ggplot2::scale_x_continuous(transform = "log2")
```

`autoplot()` shows relative WIS from `cv$score`. Raw WIS depends on the scale of
each series; relative WIS supports comparison across locations. On the log
scale, 0.5 and 2 are equally far from the reference value of 1.

Use `cv$score` for custom summaries, such as raw WIS by model and location:

```{r, cv-custom-plot}
cv$score |>
  group_by(location) |>
  arrange(wis, .by_group = TRUE) |>
  mutate(model_id_ordered = paste(model_id, location, sep = "__")) |>
  ggplot(aes(y = reorder(model_id_ordered, wis), x = wis)) +
  facet_wrap(~location, scales = "free") +
  geom_col(aes(fill = model_id), show.legend = FALSE) +
  scale_y_discrete(labels = \(x) sub("__.*$", "", x)) +
  theme_classic() +
  labs(y = "Model", x = "WIS")
```


### Forecasting

By default, `get_fcast()` combines the three best models for each location.

```{r, forecast, eval = FALSE}
fcast <- cv |> get_fcast()
```

```{r, load-fcast, include = FALSE}
fcast <- readRDS("ebola_fcast.rds")
```

```{r, show-fcast}
fcast
fcast$meta$selection
```

`autoplot()` shows the median and 50% and 95% prediction intervals.

```{r, fcast-plot, fig.cap = "Ensemble forecast by location."}
fcast |> autoplot()
```

Use `as_tibble()` to build a custom plot:

```{r, fcast-custom-plot, fig.cap = "Forecasts by model with observed values in black."}
fcast |>
  as_tibble() |>
  ggplot() +
  geom_ribbon(
    aes(x = target_end_date, ymin = lower, ymax = upper, fill = model_id),
    alpha = 0.2,
    show.legend = FALSE
  ) +
  geom_line(
    aes(x = target_end_date, y = median, colour = model_id),
    show.legend = TRUE
  ) +
  facet_wrap(~location, scales = "free") +
  geom_line(
    data = fcast$hub$oracle_output |>
      filter(target_end_date >= as.Date("2026-07-01")),
    aes(x = target_end_date, y = oracle_value),
    colour = "black"
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )
```
