---
title: "Preparing External Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing External Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Required columns

Pass external surveillance data to `check_data()` before nowcasting or
forecasting.

The data frame needs four columns:

| Column | Type | Description |
|--------|------|-------------|
| `target_end_date` | Date | Date represented by the observation |
| `observation` | numeric | Observed value |
| `location` | character | Series identifier |
| `target` | character | Target identifier |

Add `as_of` to nowcast reporting delays:

| Column | Type | Description |
|--------|------|-------------|
| `as_of` | Date | Date this version was reported |

Use one row per series and `target_end_date`, or one row per revision when
`as_of` is present. Keep one target per call. All series must use the same
reporting interval, calendar and end date, although they may start on different
dates.

`location` is the default series key. Name other key columns with `key`, for
example `check_data(df, key = c("location", "age_group"))`.

## Example

```{r, include=FALSE}
df <- data.frame(
  target_end_date = as.Date("2024-01-01") + seq(0, by = 7, length.out = 52),
  observation = rpois(52, lambda = 20),
  location = "NY",
  target = "inc hosp influenza"
)
```

```{r}
library(incast)
head(df)

checked <- check_data(df)
checked
```

Use the result directly with `get_cv()` or `get_fcast()`. Use `get_ncast()`
first when revision history is available.
