---
title: "Evaluating incast against FluSight"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluating incast against FluSight}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 8,
  dpi = 120
)

library(incast)
library(dplyr)
library(ggplot2)
```

## Overview

This vignette replays the 2025--26 US [FluSight forecasting season](https://github.com/cdcepi/FluSight-forecast-hub/tree/v1.2.0) as if `incast` had submitted a forecast each week. 

The evaluation covers:

- 7 states: Connecticut, Florida, Maryland, New Jersey, New York, North
  Carolina, and Pennsylvania;
- 27 forecast rounds from November 22, 2025, through May 23, 2026; and
- weekly incident influenza hospital admissions at horizons 0--3.

The model fitting is intentionally **not run when this vignette is built**. The
saved forecasts and score summaries are included with the package, so the result
plots below are fast and reproducible.

## Use the saved forecasts

The complete backtest is available directly as `flusight_forecasts`:

```{r saved-forecasts}
flusight_forecasts |>
  select(model_id, reference_date, horizon, location, output_type_id, value) |>
  head()
```

To create a Hubverse-format CSV for [MyRespiLens](https://myrespilens.com):

```{r copy-csv, eval = FALSE}
write.csv(flusight_forecasts, "incast.csv", row.names = FALSE)
```

The CSV contains every successfully fitted candidate. For each state and round,
the forecast from the candidate with the lowest cross-validation weighted
interval score (WIS) is duplicated with `model_id = "incast"`. Thus, `incast`
is a dynamically selected single model, not an ensemble.

## Backtest design

### Reconstruct the data available at each deadline

A FluSight round is indexed by a Saturday `reference_date`, but its submission
deadline was the preceding Wednesday. Horizon 0 predicts the week ending on the
reference date; horizons 1--3 predict the following three weeks.

For a round with reference date \(d\), the backtest uses only surveillance
records with `as_of <= d - 6 days`. For example, the November 22, 2025, round
uses data released by November 16. This reproduces the revision state available
before the November 19 submission deadline and prevents future revisions from
leaking into model selection. No reporting-delay adjustment is applied.

The following code retrieves the required revision history. It is shown for
reproducibility and is not evaluated during vignette builds.

```{r surveillance-data, eval = FALSE}
states <- c("CT", "FL", "MD", "NJ", "NY", "NC", "PA")

revision_history <- get_data(
  pathogen = "flu",
  geo_value = tolower(states),
  revisions = TRUE
)$data
```

### Select a model independently for each state and round

For each forecast round, the procedure is:

1. Use dates 52, 104, and 156 weeks before the current reference date to
   represent the same period in the three previous seasons.
2. Use that week and the next two weeks as nine cross-validation origins.
3. At each origin, fit every candidate using only earlier observations and
   score its four-week forecast against the latest values available for that
   round.
4. Select the lowest-WIS candidate separately for each state.
5. Refit every candidate to the data available for the current round and save
   all four forecast horizons. Also save the selected candidate as `incast`.

The candidates comprise the four defaults returned by `default_models()` plus
ARIMA, HHH4, Prophet, neural-network, mechanistic, and foundation models. The
complete model specification is below.

```{r models, eval = FALSE}
library(fable)
library(fable.prophet)
library(igraph)
library(incast.odin)
library(reticulate)
library(surveillance)

locations <- read.csv(
  paste0(
    "https://raw.githubusercontent.com/cdcepi/FluSight-forecast-hub/",
    "v1.2.0/auxiliary-data/locations.csv"
  )
) |>
  as_tibble() |>
  filter(abbreviation %in% states)

population <- locations |>
  mutate(population = population / sum(population)) |>
  pull(population, name = abbreviation)

neighbourhood <- graph_from_data_frame(
  rbind(
    c("CT", "NY"),
    c("MD", "PA"),
    c("NJ", "NY"),
    c("NJ", "PA"),
    c("NY", "PA")
  ),
  directed = FALSE,
  vertices = locations$abbreviation
) |>
  as_adjacency_matrix(sparse = FALSE)

annual_seasonality <- addSeason2formula(
  f = ~1,
  S = 1,
  period = round(365.25 / 7)
)

my_models <- c(
  default_models(),
  list(
    CUSTOM_ARIMA = ARIMA(observation ~ pdq(1, 1, 0)),
    ODIN2_SIR = odin_sir(observation),
    ODIN2_SEIR = odin_seir(observation),
    HHH4_AR_END = HHH4(
      observation,
      control = list(
        ar = list(f = ~1, lag = 1),
        ne = list(f = ~ -1),
        end = list(f = annual_seasonality),
        family = "NegBin1"
      ),
      population = population
    ),
    HHH4_FULL = HHH4(
      observation,
      control = list(
        ar = list(f = ~1, lag = 1),
        ne = list(f = ~1, lag = 1, normalize = FALSE),
        end = list(f = annual_seasonality),
        family = "NegBin1"
      ),
      neighbourhood = neighbourhood,
      population = population
    ),
    PROPHET = prophet(observation ~ season("year")),
    NNETAR = NNETAR(observation),
    CHRONOS = FOUNDATION(log1p(observation), "chronos"),
    TIMESFM = FOUNDATION(log1p(observation), "timesfm")
  )
)
```

The round-level implementation makes the data cutoff and seasonal
cross-validation explicit:

```{r backtest, eval = FALSE}
cv_seasons <- 2022:2024
reference_dates <- seq.Date(
  from = as.Date("2025-11-22"),
  to = as.Date("2026-05-23"),
  by = "1 week"
)

fcast_round <- function(ref_date) {
  ref_date <- as.Date(ref_date)

  round_data <- revision_history |>
    filter(as_of <= ref_date - 6L) |>
    check_data()

  last_observed_date <- round_data$window[["to"]]
  last_target_date <- ref_date + 3L * 7L
  stopifnot(last_observed_date < ref_date)

  # Forecast far enough to return horizons 0--3 even when reporting lags.
  forecast_steps <- as.integer(
    last_target_date - last_observed_date
  ) %/%
    7L

  aligned_weeks <- seq.Date(
    from = ref_date,
    by = "-52 weeks",
    length.out = length(cv_seasons) + 1L
  )[-1L]

  cv_origins <- sort(c(
    aligned_weeks,
    aligned_weeks + 7L,
    aligned_weeks + 14L
  ))

  cv <- get_cv(
    round_data,
    h = 4,
    origins = cv_origins,
    models = my_models
  )

  get_fcast(cv, h = forecast_steps, top_n = 1)$hub$model_out_tbl |>
    transmute(
      model_id = if_else(model_id == "ENSEMBLE", "incast", model_id),
      reference_date = ref_date,
      target,
      horizon = as.integer(target_end_date - ref_date) %/% 7L,
      location = locations$location[match(location, locations$abbreviation)],
      target_end_date,
      output_type,
      output_type_id,
      value = round(value)
    ) |>
    filter(horizon %in% 0:3)
}

data.table::setDTthreads(1L)

rounds <- lapply(reference_dates, function(ref_date) {
  tryCatch(
    fcast_round(ref_date),
    error = function(e) {
      warning("Round ", ref_date, " failed: ", conditionMessage(e))
      NULL
    }
  )
})

flusight_forecasts <- bind_rows(rounds)
save(
  flusight_forecasts,
  file = file.path("data", "flusight_forecasts.rda"),
  compress = "xz"
)
```

## Scoring results

The saved forecasts were scored with `hubEvals::score_model_out()` against the
FluSight v1.2.0 target data. `FluSight-ensemble` is the baseline. Lower WIS is
better, and scaled relative WIS below one indicates better performance than the
ensemble.

Only the five quantiles available for every model (0.025, 0.25, 0.5, 0.75, and
0.975) are scored. Consequently, this is a like-for-like comparison among the
models shown here, but it is not the official FluSight evaluation based on the
Hub's complete quantile grid.

We include the 3 best- and worst-performing individual FluSight models, plus the FluSight ensemble. Ranks are based on pooled scaled relative WIS across 7 states.

```{r score-overall, fig.height = 8}
plot_scores <- function(x) {
  x |>
    mutate(model_order = reorder(model_id, -wis_scaled_relative_skill)) |>
    ggplot(aes(
      x = wis_scaled_relative_skill,
      y = model_order,
      colour = model_group
    )) +
    geom_vline(xintercept = 1, colour = "grey70") +
    geom_segment(aes(xend = 1, yend = model_order), colour = "grey75") +
    geom_point(size = 2.5) +
    labs(
      x = "Scaled relative WIS (FluSight ensemble = 1)",
      y = NULL,
      colour = NULL
    ) +
    theme_classic() +
    theme(legend.position = "top")
}

plot_scores(flusight_scores)
```

Performance varies by state. In the faceted plot below, points to the left of
the vertical line outperform the FluSight ensemble within that state. The two
`odin2` candidates are omitted because they're not well parameterised.
Note `CHRONOS` and `TIMESFM` are foundation models, they may have seen the FluSight data during training.
To check that they couldn't be trained on the 2025--26 season you would need to check when the foundation models were trained on huggingface.

```{r score-by-location, fig.width = 11, fig.height = 15}
plot_scores(flusight_scores_by_location) +
  facet_wrap(~state, scales = "free_x")
```

## Recomputing the saved results

The score tables used above are stored as `flusight_scores` and
`flusight_scores_by_location`. To update the packaged forecasts and scores after
replacing `data-raw/incast.csv`, run `data-raw/FluSight.R`. The script downloads
the matching FluSight forecasts and truth and updates the `.rda` files; it does
not refit any `incast` model.

The model-fitting chunks in this vignette have `eval = FALSE` deliberately.
Run them manually only when you intend to regenerate the full backtest.
