---
title: "Forecasting connected time series with HHH4"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecasting connected time series with HHH4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

## Overview

Most `incast` models forecast each series separately. HHH4 fits all series
together, allowing recent counts in one series to help predict another.

A **unit** is one value of the series key, such as one state when
`key = "location"`. HHH4 requires a single key column and a meaningful set of
connected units.

## Model components

HHH4 models the expected count as the sum of three components:

$$
\text{expected count}
=
\text{endemic}
+
\text{autoregressive}
+
\text{neighbour}.
$$

| Component | Uses |
|---|---|
| Endemic (`end`) | Background incidence, population, and seasonality |
| Autoregressive (`ar`) | Previous counts in the same unit |
| Neighbour (`ne`) | Previous counts in connected units |

The example below uses these settings:

| Setting | Meaning |
|---|---|
| `f = ~1` | Turn a component on with one effect shared across units |
| `f = ~-1` | Turn a component off |
| `lag = 1` | Use counts from the previous reporting period |
| `S = 1` | Use one smooth annual rise and decline |
| `family = "NegBin1"` | Allow extra count variability, shared across units |
| `normalize = TRUE` | Rescale each source row of the neighbour matrix to sum to one |

`S = 1` is one **seasonal harmonic**: a sine/cosine pair that learns the timing
and size of a repeating annual cycle.

## Example

This example forecasts weekly influenza hospital admissions in Connecticut,
New Jersey, New York, and Pennsylvania. The code is not run when the vignette
is built because `get_data()` uses a live API.

### Data

```{r packages-data}
# install.packages("surveillance") # Run once if needed.
library(incast)
library(surveillance)

states <- c("ct", "nj", "ny", "pa")
flu <- get_data(pathogen = "flu", geo_value = states)

autoplot(flu)
```

HHH4 needs one count target, one series key, and aligned reporting dates. For
external data, use `check_data(data, key = "location")`.

### Connections and population

Rows of the connection matrix are sources and columns are receiving units.
Here, `1` means that two states share a border.

```{r inputs}
adjacency <- rbind(
  CT = c(CT = 0, NJ = 0, NY = 1, PA = 0),
  NJ = c(CT = 0, NJ = 0, NY = 1, PA = 1),
  NY = c(CT = 1, NJ = 1, NY = 0, PA = 1),
  PA = c(CT = 0, NJ = 1, NY = 1, PA = 0)
)

population <- c(
  CT = 3.6e6,
  NJ = 9.5e6,
  NY = 20.0e6,
  PA = 13.0e6
)
population <- population / sum(population)
```

Population is used as an offset, allowing larger states to have a larger
background count. All names must match the series key values exactly.

### Model specifications

First define one annual cycle for the endemic component:

```{r seasonality}
annual_seasonality <- addSeason2formula(
  f = ~1,
  S = 1,
  period = round(365.25 / flu$interval)
)
```

We will compare four component structures:

| Model | Endemic | Autoregressive | Neighbour |
|---|:---:|:---:|:---:|
| `HHH4_END` | on | off | off |
| `HHH4_AR_END` | on | on | off |
| `HHH4_NE_END` | on | off | on |
| `HHH4_FULL` | on | on | on |

```{r controls}
# Endemic only.
end_control <- list(
  ar = list(f = ~ -1),
  ne = list(f = ~ -1),
  end = list(f = annual_seasonality),
  family = "NegBin1"
)

# Autoregressive + endemic.
ar_end_control <- end_control
ar_end_control$ar <- list(f = ~1, lag = 1)

# Neighbour + endemic.
ne_end_control <- end_control
ne_end_control$ne <- list(f = ~1, lag = 1, normalize = TRUE)

# Full HHH4.
full_control <- ar_end_control
full_control$ne <- ne_end_control$ne
```

Create the four `incast` model specifications:

```{r models}
hhh4_models <- list(
  HHH4_END = HHH4(
    observation,
    control = end_control,
    population = population
  ),
  HHH4_AR_END = HHH4(
    observation,
    control = ar_end_control,
    population = population
  ),
  HHH4_NE_END = HHH4(
    observation,
    control = ne_end_control,
    neighbourhood = adjacency,
    population = population
  ),
  HHH4_FULL = HHH4(
    observation,
    control = full_control,
    neighbourhood = adjacency,
    population = population
  )
)
```

## Compare and forecast

Compare the HHH4 specifications with the default `incast` models using the same
forecast origins:

```{r evaluate}
models <- c(default_models(), hhh4_models)

set.seed(2026)
cv <- get_cv(
  flu,
  h = 4,
  eval_start_date = as.Date("2024-09-01"),
  models = models
)

autoplot(cv)
```

Passing `hhh4_models` to `get_fcast()` refits and forecasts all four
specifications. The default plot shows their ensemble; use `model` to inspect
each specification.

```{r forecast}
set.seed(2026)
fcast <- get_fcast(cv, models = hhh4_models)

autoplot(fcast)
autoplot(fcast, model = "HHH4_END")
autoplot(fcast, model = "HHH4_AR_END")
autoplot(fcast, model = "HHH4_NE_END")
autoplot(fcast, model = "HHH4_FULL")
```

## Add a covariate

HHH4 has a separate formula for each component, so add the covariate to the component it affects and supply its values in `control$data`. For example, school holidays may affect background incidence, so add them to the endemic component:

```{r covariate}
h <- 4L
dates <- seq(
  min(flu$data$target_end_date),
  max(flu$data$target_end_date) + h * flu$interval,
  by = flu$interval
)
school_holiday <- as.integer(format(dates, "%m") %in% c("07", "08"))

holiday_control <- full_control
holiday_control$end$f <- addSeason2formula(
  f = ~ 1 + school_holiday,
  S = 1,
  period = round(365.25 / flu$interval)
)
holiday_control$data <- list(school_holiday = school_holiday)

hhh4_models$HHH4_SCHOOL_HOLIDAY <- HHH4(
  observation,
  control = holiday_control,
  neighbourhood = adjacency,
  population = population
)
```

This simple example treats July and August as school holidays.
Time-varying covariates must cover the observed history and forecast horizon.
See `?surveillance::hhh4` for component formulas, lags, covariates, distributions,
weights, random effects, and optimizer options.

## Reference

Meyer S, Held L, Höhle M (2017). [Spatio-temporal analysis of epidemic
phenomena using the R package
surveillance](https://doi.org/10.18637/jss.v077.i11). *Journal of Statistical
Software* 77(11), 1--55.
