---
title: "incast"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{incast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 7,
  dpi = 300
)
```

## Overview

`incast` builds infectious disease forecasts in a few steps:

1. **`get_data()`** or **`check_data()`**: fetch or validate surveillance data.
2. **`get_ncast()`** *(optional)*: correct recent weeks for reporting delays.
3. **`get_cv()`** *(optional)*: evaluate candidate models by time series cross-validation.
4. **`get_fcast()`**: ensemble the best models into a forward-looking forecast.

The package follows the standard forecasting workflow described by [Hyndman & Athanasopoulos (2021)](https://otexts.com/fpp3/basic-steps.html). The overall goal is to provide public health professionals with an easily-adoptable approach to generating, evaluating forecasts, and visualising infectious disease forecasts.

To get more information about how to know whether forecasting is the best approach for your task, follow the steps in [this](forecast_planning.html) article.

```{r, setup}
library(incast)
```

## Step 1: Get data

We fetch weekly flu hospital admissions for New York and California from the [CDC NHSN](https://data.cdc.gov/Public-Health-Surveillance/Weekly-Hospital-Respiratory-Data-HRD-Metrics-by-Ju/mpgq-jmmr/about_data) via [`epidatr`](https://cmu-delphi.github.io/epidatr/).

Setting `revisions = TRUE` retrieves the full revision history (*i.e.* all past versions of the data), which is needed for nowcasting.

```{r, get_data, eval = FALSE}
# You may need a Delphi API key to run this code.
# See `?epidatr::get_api_key()` for details.
df <- get_data(pathogen = "flu", geo_value = c("ny", "ca"), revisions = TRUE)
```

You can also provide **your own data**. Just pass it through `check_data()`. See `vignette("external_data")` for formatting details.

```{r, check_data}
tail(example_data)
df <- check_data(example_data)
autoplot(df)
```

## Step 2: Nowcasting (optional)

The most recent weeks of surveillance data are almost always too low because hospitals are still filing late reports (**right truncated**). If you feed these raw counts into a forecaster, predictions will be biased downward.

`get_ncast()` estimates what the recent counts will look like once all reports arrive. With the default `max_delay = 2`, the last 2 weeks are corrected; everything before that is left untouched.

```{r, nowcast}
ncast <- get_ncast(df)
ncast
autoplot(ncast)
```

The corrected `ncast$data` contains two extra columns: `ncast_lower` and `ncast_upper` (95% CrI) for the corrected weeks. `get_fcast()` detects these automatically and uses them to propagate nowcasting uncertainty into the final forecast.

## Step 3: Forecasting

Forecasting is split into two steps:

1. **`get_cv()` (model selection)**: performs time series cross-validation on the full (median-corrected). Models are ranked by Weighted Interval Score (WIS).
2. **`get_fcast()` (final forecast)**: ensembles the best `top_n` models and generates forecasts `h` weeks ahead. When nowcast columns are available, forecasts are generated from the lower, median, and upper nowcast estimates and pooled, so prediction intervals capture both model and nowcast uncertainty.

Default models are:

* `NAIVE`: Carries the last observed value forward. A simple baseline.

* `ETS` (Exponential Smoothing): A weighted average where recent weeks matter more than older ones. Adapts to trends and seasonal patterns.

* `THETA`: Splits the data into a long-term trend and short-term fluctuations, forecasts each separately, then combines them.

* `ARIMA`: Models temporal dependence in the series using autoregressive and moving average terms. Parameters are selected automatically.

### Cross Validation

`get_cv()` performs rolling-origin [time series cross-validation](https://otexts.com/fpp3/tscv.html).

Three arguments determine how cross-validation is performed.

- **`h`**: the forecast horizon, that is, the number of reporting intervals to predict ahead.

- **`step`**: the spacing between forecast origins.
  - `step = h` (default) produces non-overlapping forecasts and is the fastest option.
  - `step < h` produces overlapping forecasts, resulting in more evaluation points but requiring more model fits.

- **`n_origins`**, **`eval_start_date`**, or **`origins`**: supply exactly one:
  - `n_origins`: the number of forecast origins.
  - `eval_start_date`: the date of the first forecast origin.
  - `origins`: explicit, possibly non-contiguous dates, such as corresponding
    weeks in previous seasons.

A forecast origin at time `d` predicts intervals `d` to `d + h - 1`, so `n_origins` origins spaced `step` intervals apart cover the last `h + (n_origins - 1) * step` reporting intervals of the series. For example, with `h = 4`, `step = 4`, and `n_origins = 3`, the evaluation period covers the last 12 intervals:

```text
Forecast 1: [d,   d+1, d+2,  d+3]
Forecast 2: [d+4, d+5, d+6,  d+7]
Forecast 3: [d+8, d+9, d+10, d+11]
```

Increasing `n_origins` provides a more reliable comparison of models, but leaves less historical data for training. Ensure that each time series contains enough observations before the first forecast origin to fit the models reliably.

The function returns an `incast_cv` object containing the cross-validation results for each model and location.

```{r, cross-validation}
cv <- get_cv(ncast, h = 4, n_origins = 16)
cv
```

Plot relative WIS by model and location using `autoplot()`. Values of `wis_relative_skill` below 1 indicate better-than-average forecasts (lower WIS), while values above 1 indicate worse-than-average forecasts (higher WIS).

```{r, plot-cv}
autoplot(cv)
```

### Forecast

`get_fcast()` ensembles the best `top_n` models from cross-validation and forecasts `h` reporting intervals ahead. `h` defaults to the horizon used in `get_cv()`.

```{r, final-forecast}
fcast <- get_fcast(cv, top_n = 2)
fcast
```

Plot the ensemble forecast with `autoplot(fcast)` (pass `model = ` to inspect any single model instead):

```{r, plot-forecast}
autoplot(fcast)
```


### Adding custom models

Any model compatible with the [`fable`](https://fable.tidyverts.org/) framework can be passed to `get_cv()`/`get_fcast()` via `models`. Epidemiological transmission models built with `odin2` can be added through [`incast.odin`](https://github.com/ACCIDDA/incast.odin). Compose with `default_models()` to keep the built-ins alongside your own:

```{r, extra-models, eval = FALSE}
library(fable)
library(fable.prophet)
library(EpiEstim)
library(projections)
library(incast.odin)

# Illustrative inputs for the joint HHH4 models. Replace the connections with
# ones appropriate to your application; rows are sources and columns recipients.
adjacency <- rbind(
  CA = c(CA = 0, NY = 1),
  NY = c(CA = 1, NY = 0)
)
population <- c(CA = 39.4e6, NY = 20.0e6)
population <- population / sum(population)
annual_seasonality <- surveillance::addSeason2formula(
  f = ~1,
  S = 1,
  period = round(365.25 / ncast$interval)
)

my_models <- c(
  default_models(),
  list(
    # Statistical models
    CUSTOM_ARIMA = ARIMA(observation ~ pdq(1, 1, 0)),

    # Epidemiological models
    EPIESTIM = EPIESTIM(
      observation,
      mean_si = 3.5,
      std_si = 2.1,
      rt_window = ncast$interval
    ),
    ODIN2_SIR = odin_sir(observation),
    ODIN2_SEIR = odin_seir(observation),
    HHH4_AR_END = HHH4(
      observation,
      control = list(
        ar = list(f = ~1, lag = 1),
        ne = list(f = ~ -1),
        end = list(f = annual_seasonality),
        family = "NegBin1"
      ),
      population = population
    ),
    HHH4_FULL = HHH4(
      observation,
      control = list(
        ar = list(f = ~1, lag = 1),
        ne = list(f = ~1, lag = 1, normalize = TRUE),
        end = list(f = annual_seasonality),
        family = "NegBin1"
      ),
      neighbourhood = adjacency,
      population = population
    ),

    # Machine learning models
    PROPHET = prophet(observation ~ season("year")),
    NNETAR = NNETAR(observation),

    # Foundation models
    CHRONOS = FOUNDATION(log(observation), "chronos"),
    TIMESFM = FOUNDATION(log(observation), "timesfm")
  )
)

cv <- get_cv(
  ncast,
  n_origins = 16,
  models = my_models
)

fcast <- get_fcast(cv, top_n = 3)
```


## Submit to RespiLens

[RespiLens](https://www.respilens.com/) is a platform for sharing respiratory disease forecasts. Drop your forecasts as `.csv` files into the myRespiLens web interface.

```{r, respilens, eval = FALSE}
write.csv(fcast$hub$model_out_tbl, "respilens.csv", row.names = FALSE)
```
