---
title: "insight.cast"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{insight.cast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 7,
  dpi = 300
)
```

## Overview

`insight.cast` builds infectious disease forecasts in a few steps:

1. `get_data()` or `check_data()` fetches or validates surveillance data.
2. `get_ncast()` optionally corrects recent weeks for reporting delays.
3. `get_cv()` optionally compares models by time-series cross-validation.
4. `get_fcast()` produces a forecast and ensemble.

See [Planning an infectious disease forecast](forecast_planning.html) before
starting a new forecasting project.

```{r, setup}
library(insight.cast)
```

## Step 1: Get data

Fetch weekly influenza hospital admissions for New York and California from the
[CDC NHSN](https://data.cdc.gov/Public-Health-Surveillance/Weekly-Hospital-Respiratory-Data-HRD-Metrics-by-Ju/mpgq-jmmr/about_data)
through [`epidatr`](https://cmu-delphi.github.io/epidatr/).

Set `revisions = TRUE` to fetch the history needed for nowcasting.

```{r, get_data, eval = FALSE}
# You may need a Delphi API key to run this code.
# See `?epidatr::get_api_key()` for details.
df <- get_data(pathogen = "flu", geo_value = c("ny", "ca"), revisions = TRUE)
```

Pass other data through `check_data()`. See `vignette("external_data")` for the
required format.

```{r, check_data}
tail(example_data)
df <- check_data(example_data)
autoplot(df)
```

## Step 2: Nowcast (optional)

Recent surveillance counts may be incomplete because reports arrive late.
Using them directly can bias forecasts downwards.

`get_ncast()` estimates their final values. By default it corrects the last two
weeks and leaves earlier observations unchanged.

```{r, nowcast}
ncast <- get_ncast(df)
ncast
autoplot(ncast)
```

For corrected weeks, `ncast$data` contains `ncast_lower` and `ncast_upper` 95%
credible interval bounds. `get_fcast()` carries this uncertainty into the
forecast.

## Step 3: Forecasting

The forecast workflow has two steps:

1. `get_cv()` evaluates models and ranks them by weighted interval score (WIS).
2. `get_fcast()` fits the models to all available data, combines the best
   `top_n` and forecasts `h` reporting intervals ahead.

The default models are:

| Model | Description |
|---|---|
| `NAIVE` | Carries the latest value forwards |
| `ETS` | Exponential smoothing |
| `THETA` | Theta method |
| `ARIMA` | Automatically selected ARIMA model |

### Cross-validation

`get_cv()` performs rolling-origin [time series cross-validation](https://otexts.com/fpp3/tscv.html).

Three arguments set the evaluation period:

- `h`: forecast horizon in reporting intervals.

- `step`: spacing between forecast origins.
  - `step = h` (default) produces non-overlapping forecasts and is the fastest option.
  - `step < h` produces more, overlapping forecasts.

- Supply one origin argument:
  - `n_origins`: the number of forecast origins.
  - `eval_start_date`: the date of the first forecast origin.
  - `origins`: explicit dates, including non-contiguous dates.

A forecast origin at time `d` predicts intervals `d` to `d + h - 1`, so `n_origins` origins spaced `step` intervals apart cover the last `h + (n_origins - 1) * step` reporting intervals of the series. For example, with `h = 4`, `step = 4`, and `n_origins = 3`, the evaluation period covers the last 12 intervals:

```text
Forecast 1: [d,   d+1, d+2,  d+3]
Forecast 2: [d+4, d+5, d+6,  d+7]
Forecast 3: [d+8, d+9, d+10, d+11]
```

More origins give a broader comparison but leave less data before the first
model fit.

The result contains forecasts and scores for each model and series.

```{r, cross-validation-fit, results = "hide"}
cv <- get_cv(ncast, h = 4, n_origins = 16)
```

```{r, cross-validation}
cv
```

Plot relative WIS by model and series. Values below 1 are better than average.

```{r, plot-cv}
autoplot(cv)
```

### Forecast

`get_fcast()` combines the best `top_n` models. Its horizon defaults to the
value used by `get_cv()`.

```{r, final-forecast-fit, results = "hide"}
fcast <- get_fcast(cv, top_n = 2)
```

```{r, final-forecast}
fcast
```

Plot the ensemble, or set `model` to inspect one model:

```{r, plot-forecast}
autoplot(fcast)
```


### Adding custom models

Pass any [`fable`](https://fable.tidyverts.org/) model through `models`.
[`insight.cast.odin`](https://github.com/ACCIDDA/insight.cast.odin) adds transmission models
built with `odin2`. Combine custom specifications with `default_models()` to
retain the defaults:

```{r, extra-models, eval = FALSE}
library(fable)
library(fable.prophet)
library(EpiEstim)
library(projections)
library(insight.cast.odin)

# Illustrative inputs for the joint HHH4 models. Replace the connections with
# ones appropriate to your application; rows are sources and columns recipients.
adjacency <- rbind(
  CA = c(CA = 0, NY = 1),
  NY = c(CA = 1, NY = 0)
)
population <- c(CA = 39.4e6, NY = 20.0e6)
population <- population / sum(population)
annual_seasonality <- surveillance::addSeason2formula(
  f = ~1,
  S = 1,
  period = round(365.25 / ncast$interval)
)

my_models <- c(
  default_models(),
  list(
    # Statistical models
    CUSTOM_ARIMA = ARIMA(observation ~ pdq(1, 1, 0)),

    # Epidemiological models
    EPIESTIM = EPIESTIM(
      observation,
      mean_si = 3.5,
      std_si = 2.1,
      rt_window = ncast$interval
    ),
    ODIN2_SIR = odin_sir(observation),
    ODIN2_SEIR = odin_seir(observation),
    HHH4_AR_END = HHH4(
      observation,
      control = list(
        ar = list(f = ~1, lag = 1),
        ne = list(f = ~ -1),
        end = list(f = annual_seasonality),
        family = "NegBin1"
      ),
      population = population
    ),
    HHH4_FULL = HHH4(
      observation,
      control = list(
        ar = list(f = ~1, lag = 1),
        ne = list(f = ~1, lag = 1, normalize = TRUE),
        end = list(f = annual_seasonality),
        family = "NegBin1"
      ),
      neighbourhood = adjacency,
      population = population
    ),

    # Machine learning models
    PROPHET = prophet(observation ~ season("year")),
    NNETAR = NNETAR(observation),

    # Foundation models
    CHRONOS = FOUNDATION(log(observation), "chronos"),
    TIMESFM = FOUNDATION(log(observation), "timesfm")
  )
)

cv <- get_cv(
  ncast,
  n_origins = 16,
  models = my_models
)

fcast <- get_fcast(cv, top_n = 3)
```


## Export to RespiLens

Save the Hubverse output as a CSV, then upload it to
[myRespiLens](https://www.respilens.com/myrespilens):

```{r, respilens, eval = FALSE}
write.csv(fcast$hub$model_out_tbl, "respilens.csv", row.names = FALSE)
```
